An efficient composition of bidirectional programs by tupling and lazy updates

### 2019-12-02 paper/2.minbigul.tex
small fixes

minBiGUL inherited from BiGUL also supports transformations $put$ and $get$ which satisfy two above laws. In addition, we restrict 'adaptive case's of BiGUL on minBiGUL. Then $put$ and $get$ satisfy one more law, \textsc{PutPut} \cite{Foster:2007:CBT:1232420.1232424}, like the following:

@@

For source/view rearrangement, BiGUL uses just one lambda expression to express how to deconstruct as well as reconstruct data. It is a kind of bijection. However, to be able to implement it in OCaml, the environment used for developing minBiGUL and solutions in the paper, we need give two functions which one is the inverse of the other. In the above syntax, $f_2 = f_1^{-1}$ and $g_2 = g_1^{-1}$.

@@

In the end of this section, let's see the minBiGUL definition of $bfoldr$ which is a putback function of an important higher-order function on lists, $foldr$:

>>>>

minBiGUL inherited from BiGUL also supports transformations $put$ and $get$ which satisfy two above laws. In addition, we restrict `adaptive case's of BiGUL on minBiGUL. Then $put$ and $get$ satisfy one more law, \textsc{PutPut} \cite{Foster:2007:CBT:1232420.1232424}, like the following:

@@

For source/view rearrangement, BiGUL uses just one lambda expression to express how to deconstruct as well as reconstruct data. It is a kind of bijection. However, to be able to implement it in OCaml, the environment used for developing minBiGUL and solutions in the paper, we need to require two functions which one is the inverse of the other. In the above syntax, $f_2 = f_1^{-1}$ and $g_2 = g_1^{-1}$.

@@

In the end of this section, let us see the minBiGUL definition of $bfoldr$ which is a putback function of an important higher-order function on lists, $foldr$:

### 2019-12-02 paper/4.xpg.tex
add/reduce spaces
small fixes

\noindent Now, let's see how we construct $pg$ recursively.

@@

            The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. This is possible when we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid.
            %We can construct a $dummy$ from the source by \textsc{GetPut} law in some cases or require programmers to give a valid view.
            Since there is no real view, this $dummy$ is necessary. The last equality changes two. First one is a change from $dummy$ to an application $construct\_dummy \ s$. This shows that we need a way to construct $dummy$ from the current source, $s$. We can require programmers to give this $construct\_dummy$ function. Second is a substitution on $s$ in $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ by $s_1$ which equals to $\putbx{bx_1}{s}{(construct\_dummy \ s)}$. The replacement of $s$ and $s_1$ is possible because of the \textsc{PutPut} law. Then, $v_3$ in the result pair $(s_3,v_3)$ equals $dummy$. So we can realize that there is no loss information when computing a composition.

@@

In the places where thrid and/or fourth argument ($s$ and $v$) are updated by applications, the computations are also accumulated in $ks$ and/or $kv$.

%Apart from the last construction, the others are quite similar to the corresponding ones of $pg$, but have some updates over the source and the view on accumulative functions $ks$ and $kv$ respectively.

Thanks to this accumulation, in $\cpginline{bx_1 \circ bx_2}$, there are only two $cpg$ calls. The first call $\cpginline{bx_1}$ requires parameter $(ks, id, s, construct\_dummy \ s)$ where $s$ and $ks$ are corresponding to the source and the update over source. Since there is no real view here, we need to construct a dummy from the source. Then the continuation updating on this dummy should be initiated as an identity function. The first $cpg$ call is assigned to a 4-tuple $(ks_1, kv_1, s_1, v_1)$. In the next assignment, a 4-tuple $(ks_2, kv_2, s_2, v_2)$ is assigned by the second $cpg$ call which uses the input as $(kv_1, kv, v_1, v)$ where $kv_1$ and $v_1$ are obtained from the result of the first assignment, and $kv$ and $v$ come from the input. It is relatively similar to the second $pg$ call assignment in $\pginline{bx_1 \circ bx_2}$. After two $cpg$ calls, a function application, $ks_1 \ s_2$, is used to produce the updated source instead of calling recursively one more time like in $\pginline{bx_1 \circ bx_2}$.

@@

    $\tab s \Leftarrow s_0; v \Leftarrow v_0; (ks, kv, s, v) \Leftarrow \cpg{bx}{\lambda \_.s}{id}{s}{v};\\
        \tab \qtab (s; v)$

@@

The problem for $cpg$ lies in redundant computation during evaluating. For keeping away such computations, we introduce an extension named $kpg$. While $cpg$ evaluate values eagerly, $kpg$ does the opposite. Every values are evaluated lazily in a computation of $kpg$. The input of $\kpginline{bx}$ is expanded to a 6-tuple $(ks, kv, ks', kv', s, v)$ where $ks$ and $kv$ keep the modification information same as $cpg$, $s$ and $v$ hold evaluated values, and $ks'$ and $kv'$ are used for lazy evaluation of actual values. The output of this function is also a 6-tuple $(ks, kv, ks', kv', s, v)$.\\
Suppose that we have a source $s_0$ and a view $v_0$. The pair of the updated source and view $(s, v)$ where $s = \putbx{bx}{s_0}{v_0}$ and $v = \getbx{bx}{s_0}$ can be obtained using $kpg$ as follows:

    $\tab s \Leftarrow s_0; v \Leftarrow v_0; (ks, kv, ks', kv', s, v) \Leftarrow \kpg{bx}{\lambda \_.s}{id}{id}{id}{s}{v};\\
        \tab \qtab (ks' \ s; kv' \ v)$

@@

In $kpg$, basically, functions for the updates will be kept in $ks'$ and $kv'$.
$\kpginline{Skip}$ and $\kpginline{Case}$ are special cases, because they require the actual values for evaluation. Therefore we evaluate the values by application of $ks' s$ and $kv' v$ and update $ks'$ and $ks'$ with the identity fuctions.
%Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$.
When evaluating $\kpginline{bx_1 \circ bx_2}$, $ks_1'$ and $s_1$ in the result of the first assignment are still redundant but they are not evaluated.

In $kpg$, we did two optimizations. First is in $\kpginline{bx_1 \times bx_2}$, the evaluation of $ks'$ and $kv'$ will be done independently in two assignments using $\kpginline{bx_1}$ and $\kpginline{bx_2}$. There may be same computations in $fst \circ ks'$ and $snd \circ ks'$ as well as $fst \circ kv'$ and $snd \circ kv'$.
To remove duplicate evaluations, we evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$.
%It is possible to evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$ to remove the redundancy like that. One more thing,
Second is in $\kpginline{Case}$. We need to evaluate $ks' \ s'$ and $kv' \ v'$ to check the $fi$ condition before returning the 6-tuple. Such evaluations can be done lazily to make programs run faster. We use the above small optimizations in our implementation.

@@

    \noindent $\xpg{bx}{s}{v} = \text{match } bx \text{ with }\\
    \tab \vert \ bx_1 \circ bx_2 \to \\
        \tabs{2} (ks_1,kv_1, ks_1', kv_1', s_1, v_1) \Leftarrow \kpg{bx_1}{\lambda \_.s}{id}{id}{id}{s}{construct\_dummy \ s};\\
        \tabs{2} (s_2, v_2) \Leftarrow \xpg{bx_2}{kv_1' \ v_1}{v};\\
        \tabs{3} (ks_1 \ s_2, v_2)\\
    \tab \vert \ \_ \to \text{similar to } put$
\end{definition}

>>>>

\noindent Now, let us see how we construct $pg$ recursively.

@@

            The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. This is possible when we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid.
            %We can construct a $dummy$ from the source by \textsc{GetPut} law in some cases or require programmers to give a valid view.
            Since there is no real view, this $dummy$ is necessary. The last equality changes two. First is a change from $dummy$ to an application $construct\_dummy \ s$. This shows that we need a way to construct $dummy$ from the current source, $s$. We can require programmers to give this $construct\_dummy$ function. Second is a substitution on $s$ in $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ by $s_1$ which equals to $\putbx{bx_1}{s}{(construct\_dummy \ s)}$. The replacement of $s$ and $s_1$ is possible because of the \textsc{PutPut} law. Then, $v_3$ in the result pair $(s_3,v_3)$ equals $dummy$. So we can realize that there is no loss information when computing a composition.

@@

In the places where third and/or fourth argument ($s$ and $v$) are updated by applications, the computations are also accumulated in $ks$ and/or $kv$.
%Apart from the last construction, the others are quite similar to the corresponding ones of $pg$, but have some updates over the source and the view on accumulative functions $ks$ and $kv$ respectively.
Thanks to this accumulation, in $\cpginline{bx_1 \circ bx_2}$, there are only two $cpg$ calls. The first call $\cpginline{bx_1}$ requires parameter $(ks, id, s, construct\_dummy \ s)$ where $s$ and $ks$ are corresponding to the source and the update over source. Since there is no real view here, we need to construct a dummy from the source same as $pg$. Then the continuation updating on this dummy should be initiated as an identity function. The first $cpg$ call is assigned to a 4-tuple $(ks_1, kv_1, s_1, v_1)$. In the next assignment, a 4-tuple $(ks_2, kv_2, s_2, v_2)$ is assigned by the second $cpg$ call which uses the input as $(kv_1, kv, v_1, v)$ where $kv_1$ and $v_1$ are obtained from the result of the first assignment, and $kv$ and $v$ come from the input. It is relatively similar to the second $pg$ call assignment in $\pginline{bx_1 \circ bx_2}$. After two $cpg$ calls, a function application, $ks_1 \ s_2$, is used to produce the updated source instead of calling recursively one more time like in $\pginline{bx_1 \circ bx_2}$.

@@

\smallvspace
    $\tab s \Leftarrow s_0; v \Leftarrow v_0; (ks, kv, s, v) \Leftarrow \cpg{bx}{\lambda \_.s}{id}{s}{v};\\
        \tab \qtab (s; v)$
\smallvspace

@@

The problem for $cpg$ lies in redundant computation during evaluating. For keeping away such computations, we introduce an extension named $kpg$. While $cpg$ evaluates values eagerly, $kpg$ does the opposite. Every values are evaluated lazily in a computation of $kpg$. The input of $\kpginline{bx}$ is expanded to a 6-tuple $(ks, kv, ks', kv', s, v)$ where $ks$ and $kv$ keep the modification information same as $cpg$, $s$ and $v$ hold evaluated values, and $ks'$ and $kv'$ are used for lazy evaluation of actual values. The output of this function is also a 6-tuple $(ks, kv, ks', kv', s, v)$.\\
Suppose that we have a source $s_0$ and a view $v_0$. The pair of the updated source and view $(s, v)$ where $s = \putbx{bx}{s_0}{v_0}$ and $v = \getbx{bx}{s_0}$ can be obtained using $kpg$ as follows:

\smallvspace
    $\tab s \Leftarrow s_0; v \Leftarrow v_0; (ks, kv, ks', kv', s, v) \Leftarrow \kpg{bx}{\lambda \_.s}{id}{id}{id}{s}{v};\\
        \tab \qtab (ks' \ s; kv' \ v)$
\smallvspace

@@

In $kpg$, basically, functions for the updates will be kept in $ks'$ and $kv'$.
$\kpginline{Skip}$ and $\kpginline{Case}$ are special cases, because they require the actual values for evaluation. Therefore we evaluate the values by application of $ks' s$ and $kv' v$ and update $ks'$ and $kv'$ with the identity functions.
%Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$.
When evaluating $\kpginline{bx_1 \circ bx_2}$, $ks_1'$ and $s_1$ in the result of the first assignment are still redundant but they are not evaluated.

Additionally we did two optimizations in $kpg$. First is in $\kpginline{bx_1 \times bx_2}$. The evaluation of $ks'$ and $kv'$ will be done independently in two assignments using $\kpginline{bx_1}$ and $\kpginline{bx_2}$. There may be same computations in $fst \circ ks'$ and $snd \circ ks'$ as well as $fst \circ kv'$ and $snd \circ kv'$.
To remove duplicate evaluations, we evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$.
%It is possible to evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$ to remove the redundancy like that. One more thing,
Second is in $\kpginline{Case}$. We need to evaluate $ks' \ s'$ and $kv' \ v'$ to check the $fi$ condition before returning the 6-tuple. Such evaluations can be done lazily to make programs run faster. We use the above small optimizations in our implementation.

@@

    \noindent $\xpg{bx}{s}{v} = \text{match } bx \text{ with }\\
    \tab \vert \ bx_1 \circ bx_2 \to \\
        \tabs{2} (ks_1,kv_1, ks_1', kv_1', s_1, v_1) \Leftarrow \kpg{bx_1}{\lambda \_.s}{id}{id}{id}{s}{construct\_dummy \ s};\\
        \tabs{2} (s_2, v_2) \Leftarrow \xpg{bx_2}{kv_1' \ v_1}{v};\\
        \tabs{3} (ks_1 \ s_2, v_2)\\
    \tab \vert \ \_ \to \text{similar to } pg$
\end{definition}

### 2019-12-02 paper/5.experiments.tex
small fixes

We have implemented and tested all methods% 
%
\footnote{All experiments on macOS 10.14.6, processor Intel Core i7 (2.6~GHz), RAM 16~GB 2400~MHz DDR4, OCaml 4.07.1. The OCaml runtime system options and garbage collection parameters are set as default. In our language, there is no type system.}$^,$%HACK!
%
\footnote{Anonymous \url{https://mega.nz/\#!yUsgyCxA!yrJfINiwtCwhmd4McYMz2xsFa-ljkgGKmMi839ut5gA}}
%
described in the previous sections.
Instead of constructing some dummies by $construct\_dummy$s, which are necessary for $pg$, $cpg$, $kpg$ and $xpg$, we replace them by the current updated views. This helps a program in the $put$ direction valid.

@@

% The first 5 cases merely use $n$ $\circ$ operators to make straight-line compositions from the same $n + 1$ \textit{pure} programs. 
The first five test cases
% (no.~1--5) 
are straight-line compositions of the same $n + 1$ non-recursive programs.
% We refer to them as `pure' compositions.
% We use the term \textit{pure} here to refer non-recursion compositions. 
The prefix \textit{lcomp} in a test name indicates that the compositions are left-associative.
% that the associate precedence of $\circ$ in a straight-line composition is left.
The suffix \textit{ldata} indicates that the input size is large. The symbol $L$ in the input column of the \textit{ldata} test cases can be expressed as a complex list as follows: $L = [T,\ldots,T] \ (n \ T\text{s}), \ T = [A,\ldots,A] \ (10 \ A\text{s}), \ A = [1,\ldots,5]$. These structures are only intended to generate test data that is large enough for 
% observing 
measuring
% experimental 
results.

@@

$\putbx{ptail}{s}{v}$ accepts a source list $s$ and a view list $v$ to produce a new list by replacing the tail of $s$ with $v$. $\getbx{ptail}{s}$ returns the tail of the source list $s$. The compositions of many $ptail$, in the \textit{put} direction, replace a part of the tail of the source list by the view list, in the \textit{get} direction, returns the such tail from the the source.

@@

\textit{breverse} is defined in terms of \textit{bfoldr}, in the \textit{put} direction, produces a reverse of the view list if the source list has a same length with the view, in the \textit{get} direction, produce a reverse of the source list. Note that compositions is by the recursions of $breverse$ and 
the number of compositions are dynamically determined by the length of the source list.

@@

Figure \ref{fig:result} illustrates the evaluation times of $put$ in minBiGUL, $put_m$ in minBiGUL$_m$ and $xpg$ for 7 test cases. We simply ignore the experimental results of $pg$, $cpg$ and $kpg$ since they are shown the slowness compared to $xpg$. For left associative compositions, $put$ works poorly because of the number of reevaluated $get$s.
%The figure shows $put_m$ and $xpg$ are better than $put$ in some cases.
The left part of the figure contains tests using not-large inputs, and we see that $put_m$ is the fastest method for not-large inputs. However if the input size is large enough, in the cases of the right part of the figure, $put_m$ will be slower quickly due to times for manipulating data in the table. At that time, $xpg$ proved to be the most effective method.

>>>>

We have implemented and tested all methods% 
%
\footnote{All experiments on macOS 10.14.6, processor Intel Core i7 (2.6~GHz), RAM 16~GB 2400~MHz DDR4, OCaml 4.07.1. The OCaml runtime system options and garbage collection parameters are set as default. In our target language, there is no type system.}$^,$%HACK!
%
\footnote{Anonymous \url{https://mega.nz/\#!yUsgyCxA!yrJfINiwtCwhmd4McYMz2xsFa-ljkgGKmMi839ut5gA}}
%
described in the previous sections.
Instead of constructing some dummies by $construct\_dummy$s, which are necessary for $pg$, $cpg$, $kpg$ and $xpg$, we replace them by the current updated views. This helps a program in the $put$ direction valid.

@@

% The first 5 cases merely use $n$ $\circ$ operators to make straight-line compositions from the same $n + 1$ \textit{pure} programs. 
The first 5 test cases
% (no.~1--5) 
are straight-line compositions of the same $n + 1$ non-recursive programs.
% We refer to them as `pure' compositions.
% We use the term \textit{pure} here to refer non-recursion compositions. 
The prefix \textit{lcomp} in a test name indicates that the compositions are left-associative.
% that the associate precedence of $\circ$ in a straight-line composition is left.
The suffix \textit{ldata} indicates that the input size is large. The symbol $L$ in the input column of the \textit{ldata} test cases can be expressed as a complex list as follows: $L = [T,\ldots,T] \ (n \ T\text{s}), \ T = [A,\ldots,A] \ (10 \ A\text{s}), \ A = [1,\ldots,5]$. These structures are only intended to generate test data that is large enough for 
% observing 
measuring
% experimental 
results.

@@

$\putbx{ptail}{s}{v}$ accepts a source list $s$ and a view list $v$ to produce a new list by replacing the tail of $s$ with $v$. $\getbx{ptail}{s}$ returns the tail of the source list $s$. The compositions of many $ptail$, in the \textit{put} direction, replace a part of the tail of the source list by the view list, in the \textit{get} direction, returns the such tail from the source.

@@

\textit{breverse} is defined in terms of \textit{bfoldr}, in the \textit{put} direction, produces a reverse of the view list if the source list has a same length with the view, in the \textit{get} direction, produce a reverse of the source list. Note that compositions are by the recursions of $breverse$ and 
the number of compositions are dynamically determined by the length of the source list.

@@

Figure \ref{fig:result} illustrates the evaluation time of $put$ in minBiGUL, $put_m$ in minBiGUL$_m$ and $xpg$ for 7 test cases. We simply ignore the experimental results of $pg$, $cpg$ and $kpg$ since they are shown the slowness compared to $xpg$. For left associative compositions, $put$ works poorly because of the number of reevaluated $get$s.
%The figure shows $put_m$ and $xpg$ are better than $put$ in some cases.
The left part of the figure contains tests using not-large inputs, and we see that $put_m$ is the fastest method for them. However if the input size is large enough, in the cases of the right part of the figure, $put_m$ will be slower quickly due to time for manipulating data in the table. At that time, $xpg$ is proved to be the most effective method.

### 2019-12-02 paper/7.conclusion.tex
small fixes

In this paper, we focus on efficiency of composition of BX programs.
To achieve fast evaluation, we introduce two different methods using memoization.
From the experimental results of left associative BX composition programs, we know that $put_m$ is fastest method if input size is not large and $xpg$ is fastest for other left associative programs.
This shows that if programmer choose one method based on the their BX programs and inputs, they can use efficient evaluation.

@@

We will continue our work on the following three points.
First is overcomming our limitation in $xpg$. In $xpg$, there are two limitations: only for very-well-behaved programs and restriction for $\pginline{Case}$ in Section~\ref{sect:pg}. To be a practical evlauation method, extention of the target language is needed. After extention of the language, we can evaluate more various programs for experimental results.

Second is introducing an automatic analysis about BX programs and inputs to choose the best evaluation method.
Currently, programmers have to choose the evaluation method by themselves based on their BX programs and inputs. If this is achieved, we can reduce programmer's burden.

Third is introducing type systems to datatype of sources and view. If we introduce this, we can avoid runtime errors like statically-typed functional languages. We need to investigate more about $construct\_dummy$, because current definition that is used in Section~\ref{sect:experiments} will cause type errors.

>>>>

In this paper, we focus on efficiency of composition of BX programs.
To achieve fast evaluation, we introduce two different methods using memoization.
From the experimental results of left associative BX composition programs, we know that $put_m$ is fastest method if input size is not large and $xpg$ is fastest for other left associative programs.
This shows that if programmers choose one method based on their BX programs and inputs, they can use an efficient evaluation method.

@@

We will continue our work on the following three points.
First is overcoming our limitation in $xpg$. In $xpg$, there are two limitations: only for very-well-behaved programs and restriction for $\pginline{Case}$ in Section~\ref{sect:pg}. To be a practical evaluation method, an extension of the target language is needed. After extension of the language, we can evaluate more various programs for experimental results.
Second is introducing an automatic analysis about BX programs and inputs to choose the best evaluation method.
Currently, programmers have to choose the evaluation method by themselves based on their BX programs and inputs. If this analysis is achieved, we can reduce programmers' burden.
Third is introducing a type system to our target language, especially the datatype of sources and view. If we introduce this, we can avoid runtime errors like statically-typed functional languages. For this, we need to investigate more about $construct\_dummy$, because the current definition that is used in our experiments %Section~\ref{sect:experiments}
will cause type errors.
